# One-way within-subjects Analysis of Variance (Repeated one-way ANOVA) in R {#sec-anovawithin}

::: {.callout-tip appearance="simple"}
## Get the Code

You can download the complete R script from here:

<a href="https://osf.io/download/r9w2m/" download="psy400_4.R"> <button type="button" class="btn btn-primary"> <i class="bi bi-download"></i> Download R Script </button> </a>
:::

::: {.callout-tip appearance="simple"}
## Get the Data symptoms

You can download the Excel dataset **symptoms** from here:

<a href="https://osf.io/download/ceg9b/" download="symptoms.xlsx"> <button type="button" class="btn btn-success"> <i class="bi bi-file-earmark-excel"></i> Download Excel Data </button> </a>
:::

<br>

First, we load the following R packages in our R session:

```{r}
#| warning: false 

# Required packages for this R script
library(tidyverse)
library(jmv)
library(afex)
library(emmeans)
library(dlookr)
library(readxl)
```

<br>

## Introduction

The one-way repeated measures analysis of variance (also known as a within-subjects ANOVA) is an extension of the paired t-test designed to assess whether there are significant differences in the means of three or more related groups, such as comparing the difference between three or more time points.

Although, ANOVA can detect whether there are mean differences between the related groups, it does not identify which groups are different from the others. Therefore, after an ANOVA test concludes that at least one difference exists between the related groups (omnibus analysis), we should perform statistical tests that account for the number of comparisons (post hoc tests). Some of the most commonly used post hoc tests include the Tukey test, Bonferroni correction, and Holm test.

## Research question and anova assumptions

A study investigated the effect of a drug on symptom intensity in a specific disease, assessing eight patients before (two pre-test baseline measures) and after treatment (three post-test measures). The primary question is whether the mean symptom intensity score changed over time.

### Hypothesis testing for the one-way ANOVA test

::: {.callout-note icon="false"}
## Null hypothesis and alternative hypothesis

-   $H_0$: all related group means are equal (the means of symptom intensity score in the five time points are equal: $\mu_{1} = \mu_{2} = \mu_{3} = \mu_{4} = \mu_{5}$)
-   $H_1$: at least one group mean differs from the others (there is at least one group with mean symptom intensity score different from the others)
:::

### Assumptions

**A. Normality:** Scores are approximately normally distributed at each time point. To increase the number of observations used in assessing normality, we often examine the residuals from the ANOVA model for the entire dataset using a normal quantile plot of the standardized residuals (Normal Q-Q plot).

**B. Sphericity:** The variances of differences between all pairs of time points are equal (Mauchly's test).

## Import dataset and process the data variables

In this example, symptom intensity was evaluated on a twelve-point scale with higher values indicating more intense symptoms. Two pre-treatment baseline measurements were made, the first a month prior to treatment and the second one week prior to treatment. Post-treatments measures of symptom intensity were made after one week, one month, and one year following administration of the drug. In this simplified example, we have eight patients in the study.

If the dataset *symptoms* is saved in the subfolder named “data” of our RStudio project, we can read it with the `read_excel()` function from `readxl` package as follows:

```{r}
#| warning: false 

dat <- read_excel("data/symptoms.xlsx")
dat
```

<br>

The dataset has 8 patients (rows) and is organized in a “wide” format. We need to reshape the data in long format using the `pivot_longer()`.

```{r}
dat_long <- dat |> 
  pivot_longer(
    cols = c(pre1, pre2, post1, post2, post3),
    names_to = "timepoint",
    values_to = "score"
  ) |> 
  mutate(
    subid = factor(subid),
    timepoint = factor(timepoint, 
                       levels = c("pre1", "pre2", "post1", "post2", "post3"),
                       labels = c("prior1m", "prior1w", "post1w", "post1m", "post1y"))
    )

dat_long
```

***Explanation***

The data initially starts in a wide format, where each time point is stored in its own column (e.g., pre1, pre2, post1, post2, post3). The `pivot_longer()` function reshapes the data into a long format by stacking these columns into two new variables, with names that we specify (e.g., timepoint and score):

-   **timepoint:** Indicates when the measurement was taken.
-   **score:** Contains the corresponding symptom value.

Then, after reshaping to long format, we use `mutate()` to convert the `subid` and `timepoint` variables into factors (`<fct>`) by applying the `factor()` function within `mutate()`.

<br>

## Descriptive statistics and plots

We use functions from the ***dplyr*** and ***dlookr*** packages to obtain summary statistics by study group:

```{r}
# descriptive statistics using the package dlookr
dat_long |>
  group_by(timepoint) |>
  describe(score) |>
  select(timepoint, n, na, mean, sd, p25, p50, p75, skewness, kurtosis)

```

<br>

A box plot is useful for visualizing the central tendency and dispersion of numerical data, especially when comparing distributions across multiple groups.

```{r}
ggplot(dat_long, aes(x = timepoint, y = score)) +
  geom_boxplot(fill = "#B0C4DE", width = 0.4) +
  stat_summary(fun = mean, geom = "point", 
               shape = 15, size = 2.5, color = "deeppink")
```

The graph shows stable symptom intensity at the two measurements prior the treatment, followed by a clear, progressive reduction across the post-treatment time points.

## Define the ANOVA model

First, we define the ANOVA model using the `aov()` function:

```{r}
anova_model <- aov_car(score ~ timepoint + Error(subid/timepoint), data = dat_long)
```

It creates a model object that stores the mathematical association between time points and scores.

## Check ANOVA Assumptions

### Normality test and Q-Q plot

We can conduct a Shapiro-Wilk test to assess the normality of the standardized residuals from the ANOVA model:

```{r}
# Get raw residuals
resids <- residuals(anova_model)

# Get standardized residuals
std_resids <- scale(resids)

# Run Shapiro-Wilk Test
shapiro.test(std_resids)
```

The Shapiro-Wilk test of normality suggests normal distributions (p=0.23 \> 0.05; $H_o$ is not rejected).

<br>

We can also examine the normality assumption using a normal quantile plot (Q-Q plot) of the standardized residuals:

```{r}
# Q-Q plot
qqnorm(std_resids)
# add a reference line
qqline(std_resids, col = "red", lwd = 2)
```

The data points mostly fall along the diagonal line, indicating that the residuals are approximately normally distributed.

<br>

### Sphericity assumption

This assumption is usually checked with the **Mauchly's sphericity test**, where null hypothesis states that the variances of the differences are equal.

```{r}
summary(anova_model)$sphericity.tests
```

In our example, the assumption of sphericity has not been met (p = 0.043). In this case, we have to correct the degrees of freedom in repeated ANOVA analysis.

There are two correction options $\epsilon$ (epsilon) available: Greenhouse-Geisser (GG eps), or Huynh-Feldt (HF eps). The general recommendation is to use the Greenhouse-Geisser $\epsilon$ correction when it is less than 0.75; otherwise, we should use the Huynh-Feldt $\epsilon$ correction.

```{r}
summary(anova_model)$pval.adjustments
```

As the *GG eps* value is less than 0.75, we use the Greenhouse-Geisser adjustment of 0.488.

The corrected degrees of freedom are:

$df_1*GG \ eps=4*0.488=1.95$

and

$df_2*HF \ eps=28*0.488=13.66$.

## Perform ANOVA test (omnibus analysis)

We use ***jmv*** package to perform a Repeated Measures ANOVA because of its precise partitioning of variance. Unlike standard R outputs, **`anovaRM()`** function explicitly separates the total variance into two distinct tables: Within-Subjects Effects and Between-Subjects Effects.

```{r}
results <- jmv::anovaRM(
  data = dat,    # <1>
  rm = list(
    list(
      label = "timepoint",
      levels = c("pre1", "pre2", "post1", "post2", "post3")
    )
  ),
  rmCells = list(
    list(measure = "pre1", cell = "pre1"),
    list(measure = "pre2", cell = "pre2"),
    list(measure = "post1", cell = "post1"),
    list(measure = "post2", cell = "post2"),
    list(measure = "post3", cell = "post3")
  ),
  spherCorr = "GG"   # <2>
)

```

1.  The `anovaRM()` function needs wide format data.
2.  We use the Greenhouse-Geisser correction as it is less than 0.75.

<br>

-   **Within-subjects variance table**

```{r}
#| column: page-right

results$rmTable  # Within Subjects
```

<br>

::: panel-tabset
## F-distribution

```{r}
#| echo: false
#| message: false
#| warning: false


# Data from GG-corrected analysis
df1_gg <- 1.95
df2_gg <- 13.66
f_obs  <- 18.62

# Generate data for the main peak (0 to 10)
x_vals <- seq(0, 10, length.out = 500)
df_plot <- data.frame(x = x_vals, y = df(x_vals, df1_gg, df2_gg))

# Create the Overview Plot
ggplot(df_plot, aes(x = x, y = y)) +
  geom_line(color = "steelblue", size = 1.2) +
  annotate("segment", x = 8, xend = 9.5, y = 0.05, yend = 0.05, 
           arrow = arrow(length = unit(0.3, "cm")), color = "red") +
  annotate("text", x = 7.2, y = 0.15, label = paste("Observed F = 18.62\np < 0.001"), color = "red") +

  labs(title = "F-Distribution (1.95, 13.66)",
       subtitle = paste("The observed F-statistic (18.62) is far off-screen to the right. Click the Zoom In tab to see the tail."),
       x = "F-statistic",
       y = "Density") +
  
  theme_minimal(base_size = 20) +
  theme(
    plot.margin = margin(10, 10, 40, 10),
    panel.grid.minor = element_blank(),
    plot.background = element_rect(fill = "#fffafa", color = NA)
  )

```

## Zoom In tail

```{r}
#| echo: false
#| message: false
#| warning: false

# Data from GG-corrected analysis
df1_gg <- 1.95
df2_gg <- 13.66
f_obs  <- 18.62

# Calculate the exact height (y) of the distribution curve at our observed F
# This ensures the dashed line stops exactly at the blue line
f_height <- df(f_obs, df1_gg, df2_gg)

# Extreme Tail Zoom Plot
ggplot(data.frame(x = c(12, 22)), aes(x)) +
  # Draw the F-distribution curve
  stat_function(fun = df, args = list(df1 = df1_gg, df2 = df2_gg), 
                color = "steelblue", size = 1.0) +
  
  # Shade the p-value area
  stat_function(fun = df, args = list(df1 = df1_gg, df2 = df2_gg),
                xlim = c(f_obs, 22), geom = "area", fill = "#d35", alpha = 0.5) +
  
  # FIXED LINE: Segment starting from the axis (0) up to the curve height
  annotate("segment", x = f_obs, xend = f_obs, y = 0, yend = f_height,
           linetype = "dashed", color = "red", size = 1) +
  
  # F-VALUE ON THE AXIS
  annotate("text", x = f_obs, y = -0.00001, 
           label = paste("F =", round(f_obs, 2)), 
           color = "black", fontface = "bold", vjust = 1.5, size = 6.5) +
  
  # P-VALUE IN THE SHADED AREA
  annotate("text", x = f_obs + 1.5, y = 0.000009, 
           label = "p = 0.00013", 
           color = "red", fontface = "bold", size = 6.5) +
  
  # THE EXTREME ZOOM
  coord_cartesian(xlim = c(12, 22), ylim = c(0, 0.0002), clip = "off") +
  
  labs(title = paste("F-Distribution (", df1_gg, ",", df2_gg, "): Extreme Tail Zoom"),
       x = "F-statistic",
       y = "Probability Density)") +
  theme_minimal(base_size = 20) +
  # Light Pink Aesthetic
  theme(plot.margin = margin(10, 10, 30, 10), 
    panel.grid.minor = element_blank()
  )
```
:::

As we can see, the within-subjects variable (timepoint) was statistically significant under the Greenhouse-Geisser correction, F(1.95, 13.66) = 18.62, p \< 0.001.

A significant one-way ANOVA is generally followed up by post-hoc tests to perform multiple pairwise comparisons between groups, while controlling for the overall error rate.

<br>

-   **Between-subjects variance table**

```{r}
results$bsTable  # Between Subjects
```

<br>

## Conduct Post hoc tests

To follow up on the significant main effect of time, pairwise comparisons were conducted using the Holm–Bonferroni adjustment. This procedure controls the family-wise error rate while being less conservative than the standard Bonferroni correction, thereby providing greater statistical power and reducing the likelihood of Type II errors.

```{r}
post_hocs <- emmeans(anova_model, ~ timepoint)
pairs(post_hocs, adjust = "holm")
```

For example, the mean difference between prior1m and post1m is: 9.12 - 5.25 = 3.87, which is significant (adjusted p = 0.021).

 

::: content-box-green
**Interpretation**

Based on Greenhouse-Geisser correction for violation of sphericity, a one-way within-subjects ANOVA revealed a significant difference in the pre-test and post-test means, F(1.95, 13.66) = 18.62, p \< 0.001. Pairwise comparisons using a Holm-Bonferroni correction to maintain an alpha level of 0.05 revealed that intensity of symptoms remained constant from the two pre-test baseline measures, significantly decreased after a week following drug therapy, and further significantly decreased after one month. Symptom intensity did not significantly differ from that level at the end of a year.
:::
