# Basic data structures in R {#sec-introduction}

::: {.callout-tip appearance="simple"}
## Get the Code

You can download the complete R script here:

<a href="https://osf.io/download/nzv38/" download="psy400_0.R"> <button type="button" class="btn btn-primary"> <i class="bi bi-download"></i> Download R Script </button> </a>
:::


## Atomic Vectors

-   Numeric vectors

```{r}
# numeric vector
a <- c(2, 5, 4)         # c stands for `concatenate' function
a
class(a)
```

<br>

```{r}
# quickly create a vector of consecutive numbers with the colon operator
a <- 20:100
a
```

<br>

```{r}
# For more general sequences we can use the `seq()` function
a <- seq(from = 20, to = 100, by = 0.5)  # explicitly named arguments
a
```

<br>

```{r}
# positional matching: match the arguments in the order that they appeared
seq(20, 100, 0.5)
```

<br>

-   Logical vector

```{r}
b <- c(TRUE, FALSE, TRUE, TRUE, TRUE, TRUE, TRUE, FALSE)
b
class(b)
```

<br>

-   Character/string vector

```{r}
d <- c("Male", "Male", "Male", "Female", "Female", "Female")
d

class(d)
```

<br>

-   Factor vector

```{r}
# a factor variable is a special kind of character vector

f <- as.factor(d)  
f  

class(f)
```

<br>

-   Date vector

```{r}
# also dates are objects with their own class

g <- as.Date(c("2020-02-18", "2020-05-27"))
g
class(g)
```

## Matrix (2-dimensional atomic vector)

```{r}
a <- c(2, 1, 4, 5, 3, 7) # numeric vector with six elements


# Matrices are constructed column-wise starting in the “upper left” corner 
# and running down the columns
A <- matrix(a, nrow=2, ncol=3)
A
```

In R, we can access rows, columns, or individual elements of a matrix using subscripts within brackets.

```{r}
# select the second row
A[2, ]
```

```{r}
# select the 2nd column
A[ , 2]
```

```{r}
# select the element in the 2nd row, 3rd column
A[2, 3]

```

## Data frame (generic vector or list)

```{r}
# build data frames using vectors

patientID <- c(1, 2, 3, 4, 5, 6, 7, 8)

age <- c(25, 30, 28, 22, 31, 45, 37, 43)

weight <- c(94, 83, 71, 87, 94, 73, 89, 74)

diabetes <- c("Type 1", "Type 2", "Type 1", "Type 1",
              "Type 2", "Type 1", "Type 1", "Type 2")

```

<br>

```{r}
# create the dataframe
dat <- data.frame(patientID, age, weight, diabetes) # insert all the variables
dat
```

```{r}
typeof(dat)
class(dat)
```
