# Non-parametric statistical tests in R {#sec-nonparamteric}

::: {.callout-tip appearance="simple"}
## Get the Code

You can download the complete R script from here:

<a href="https://osf.io/download/cy62s/" download="psy400_2.R"> <button type="button" class="btn btn-primary"> <i class="bi bi-download"></i> Download R Script </button> </a>
:::

::: {.callout-tip appearance="simple"}
## Get the Data maths

You can download the Excel dataset **maths** from here:

<a href="https://osf.io/download/pw8q9/" download="maths.xlsx"> <button type="button" class="btn btn-success"> <i class="bi bi-file-earmark-excel"></i> Download Excel Data </button> </a>
:::

::: {.callout-tip appearance="simple"}
## Get the Data boxbreathing

You can download the Excel dataset **boxbreathing** from here:

<a href="https://osf.io/download/7fmbu/" download="boxbreathing.xlsx"> <button type="button" class="btn btn-success"> <i class="bi bi-file-earmark-excel"></i> Download Excel Data </button> </a>
:::

<br>

First, we load the following R packages in our R session:

```{r}
#| warning: false 

# Required packages for this R script
library(tidyverse)
library(ggpubr)
library(dlookr)
library(readxl)
```

<br>

## INDEPENDENT SAMPLES

In this lecture, the Wilcoxon-Mann-Whitney (WMW) test, also known as
the Mann-Whitney U test or Wilcoxon Rank Sum test, is presented as a nonparametric method for comparing two independent samples. It is often used as an alternative to the independent samples t-test, particularly when the assumption of normality is violated or when dealing with small sample sizes.



### Research question

A high school is pilot-testing a new curriculum for mathematics. To ensure the curriculum is equitable, the department head has collected the final exam scores from a small sample of students to test for significant differences in performance between male and female students.


### Import dataset

If the dataset *maths* is saved in the subfolder named “data” of our RStudio project, we can read it with the `read_excel()` function from `readxl` package as follows:

```{r}
#| warning: false 

dat_maths <- read_excel("data/maths.xlsx")
dat_maths
```

<br>


The dataset has 17 students (rows) and is organized in a “long” format. It includes two variables (columns): the character (`<chr>`) sex variable (independent variable) and the numeric (`<dbl>`) score variable (dependent variable).




### Descriptive statistics and Q-Q plots


```{r}
# descriptive statistics using the package dlookr
dat_maths |>
  group_by(sex) |>
  describe(score) |>
  select(sex, n, na, mean, sd, p25, p50, p75, skewness, kurtosis)

```

<br>


```{r}
# Q_Q plots using the package ggpubr
ggqqplot(dat_maths, "score", color = "sex", conf.int = FALSE, facet.by = "sex")
```




### Wilcoxon-Mann-Whitney test

```{r}
wilcox.test(score ~ sex, data = dat_maths)
```

The results indicated no statistically significant difference ($W = 43, p = 0.54$).


<br>


## DEPENDENT SAMPLES

The Wilcoxon Signed-Rank test is a non-parametric statistical test used
to compare two groups of paired observations. It assesses whether the paired differences are centered around zero by considering both
the magnitudes and signs of the differences, using their signed ranks. This test is often preferred when the paired differences are not normally distributed or when sample sizes are small.



### Research question

Researchers aim to determine whether a specific breathing technique, Box Breathing, can reduce individuals’ heart rate, measured in beats per minute (BPM), immediately before they deliver a public speech. They measure the heart rates of 10 participants during a baseline stress period and again after the participants practice the breathing technique for five minutes.



### Import dataset

If the dataset *boxbreathing* is saved in the subfolder named “data” of our RStudio project, we can read it with the `read_excel()` function from `readxl` package as follows:

```{r}
dat_BPM <- read_excel("data/boxbreathing.xlsx")
dat_BPM
```

The dataset has 10 participants and is organized in a “wide” format. It includes two columns with numeric (`<dbl>`): the baseline BPM and after the participants practice the breathing technique for five minutes.


<br>

First, we use the `mutate()` function to calculate the difference
between the boxbreath and the baseline measurements of the BPM, as follows:

```{r}
dat_BPM <- dat_BPM |>
  mutate(dif_BPM = boxbreath_BPM - baseline_BPM)

dat_BPM
```



### Descriptive statistics and Q-Q plot


```{r}
# descriptive statistics using the package dlookr

dat_BPM |>
  dlookr::describe(dif_BPM, boxbreath_BPM, baseline_BPM) |>
  select(described_variables, n, mean, sd, p25, p50, p75, skewness, kurtosis)

```

<br>

```{r}
# Q_Q plots using the package ggpubr
ggqqplot(dat_BPM, "dif_BPM", conf.int = FALSE)

```



### Wilcoxon Signed-Rank Test

```{r}
# Wilcoxon-Mann-Whitney test (non-parametric test)
wilcox.test(dat_BPM$boxbreath_BPM, dat_BPM$baseline_BPM, paired = TRUE)
```

<br>

```{r}
# or use the differences
wilcox.test(dat_BPM$dif_BPM)
```

The results indicated a statistically significant decrease in BPM ($V = 3, p = 0.009$). The median BPM (85) after the intervention was significantly lower than the baseline (94.5), suggesting the technique is an effective method for relaxing.
