# Data processing, transformation and ploting in R {#sec-process}

::: {.callout-tip appearance="simple"}
## Get the Code

You can download the complete R script from here:

<a href="https://osf.io/download/pyc9q/" download="psy400_1.R"> <button type="button" class="btn btn-primary"> <i class="bi bi-download"></i> Download R Script </button> </a>
:::

::: {.callout-tip appearance="simple"}
## Get the Data Shock index

You can download the Excel dataset **shock_index** from here:

<a href="https://osf.io/download/enj2h/" download="shock_index.xlsx"> <button type="button" class="btn btn-success"> <i class="bi bi-file-earmark-excel"></i> Download Excel Data </button> </a>
:::

::: {.callout-tip appearance="simple"}
## Get the Data cholesterol

You can download the Excel dataset **cholesterol** from here:

<a href="https://osf.io/download/fv26s/" download="cholesterol.xlsx"> <button type="button" class="btn btn-success"> <i class="bi bi-file-earmark-excel"></i> Download Excel Data </button> </a>
:::

We will use the `tidyverse` package, a comprehensive suite of R packages designed for data manipulation, transformation, and visualization tasks, all of which work together in harmony.

```{r}
#| warning: false 

# Import tidyverse

library(tidyverse)
```

<br>

## Import dataset

If the dataset is saved in the subfolder named “data” of our RStudio project, we can read it with the `read_excel()` function from `readxl` package as follows:

```{r}
#| warning: false 

library(readxl)
dat <- read_excel("data/shock_index.xlsx")
dat
```

<br>

Choosing informative column names can significantly enhance the clarity and interpretability of our dataset. For instance, renaming `HR` to `Heart.Rate` offers a more intuitive and informative label for the heart rate variable.

```{r}
# rename HR to Heart.Rate
dat <- rename(dat, Heart.Rate = HR)
dat

```

<br>

It’s crucial to ensure that the column names are consistent with the naming conventions in R. This can be achieved by using the `clean_names()` function from the `janitor` package, which standardizes column names by converting them to lowercase, removing special characters, and replacing spaces with underscores.

```{r}
#| warning: false 

# load the janitor package
library(janitor)

# clean names
dat <- clean_names(dat)
dat
```

## Converting to the appropriate data type

We might have noticed that the categorical variable sex is coded as F (for females) and M (for males), so it is recognized of character type. We can use the `factor()` function to encode a variable as a factor:

```{r}
# convert sex to a factor variable with levels female, male
dat$sex <- factor(dat$sex, levels = c("F", "M"), labels = c("female", "male"))
dat
```

## Transforming data

### Subsetting observations (rows)

Subsetting observations (rows) is a common data manipulation task in data analysis, allowing us to focus on specific subsets of data that meet particular criteria. This process involves selecting rows from a dataset based on conditions we specify.

- **Basic way**

Just as we use square brackets `[row, column]` to select elements from matrices, we apply the same approach to select elements from a data frame. By applying logical conditions, such as greater than (\>), less than (\<), equal to (==), or combinations of these, we can filter observations that meet specific criteria. For example, we can filter the data to find patients whose age is greater than 55, as follows:

```{r}
# select rows where age > 55
dat1 <- dat[dat$age > 55, ]
dat1

```

Similarly, we can use the `which()` function that returns the indices where the condition `age > 55` is `TRUE`. However, in this case, the missing values are treated as `FALSE`, resulting in the exclusion of rows where age in the original data is NA.

```{r}
dat2 <- dat[which(dat$age > 55), ]
dat2
```

<br>

If our objective is to analyze data for female patients aged over 55, we can create a subset by incorporating the logical `&` operator to combine the conditions for age and sex.

```{r}
# select rows where age > 55 and sex is female
dat3 <- dat[which(dat$age > 55 & dat$sex == "female"), ]
dat3
```

<br>


- **Tidyverse way**

The `dplyr` package also provides the `filter()` function for selecting observations. Selecting rows where age is greater than 55:

```{r}
# select rows using filter

dat4 <- filter(dat, age > 55)
dat4
```

Selecting rows where age is greater than 55 and patient is female:

```{r}
dat5 <- filter(dat, age > 55, sex == "female")
dat5
```

### Subsetting variables (columns)

When working with data frames, selecting specific variables (columns) for analysis is a common task. For instance, suppose we want to select only the `age`, and `heart_rate` variables from the data frame.


- **Basic way**

Subsetting variables can be achieved by indexing the variables’ names within `[ ]`, as follows:

```{r}
dat6 <- dat[ , c("age" , "heart_rate")]
dat6
```

By specifying the column indices, such as `c(2, 4)`, within brackets, we can select the data in columns 2, and 4 as follows:

```{r}
dat7 <- dat[ , c(2, 4)]
dat7
```


<br>

- **Tidyverse way**

In `select()` function we pass the data frame first and then the variables separated by commas. Quotes are optional for column names that don’t contain special characters or spaces.

```{r}
dat8 <- select(dat, age, heart_rate)
dat8
```

Or we can exclude the variables `sex` and `sbp`.

```{r}
dat9 <- select(dat, -c(sex, sbp))
dat9
```

### Creating new variables

Data analysis often involves creating new variables within data frames based on existing variables. This data transformation process involves adding new columns to a data frame derived from the values of one or more existing columns.

Suppose we want to compute a new variable representing the ratio of heart rate to systolic blood pressure (HR/sbp), which we will call shock_index. We can achieve this by using the `mutate()` function from `dplyr` package, as follows:

```{r}
dat10 <- mutate(dat, shock_index = heart_rate / sbp)

dat10
```

Our next step involves categorizing the newly created shock_index variable based on the defined normal range (0.5 to 0.7 beats/mmHg min). This can be done using the `case_when()` function from the `dplyr` package.

```{r}
# convert shock_index into a categorical variable 
# (<0.5 with low, <= 0.7 with normal, > 0.7 high)

dat11 <- mutate(dat10,
                shock_index1 = case_when(shock_index < 0.5 ~ "low",
                                         shock_index <= 0.7 ~ "normal",
                                         TRUE ~ "high"))
dat11
```

However, it’s important to keep in mind that transforming continuous variables into categories may simplify interpretation, but it often results in information loss.

## Using the native pipe operator in a sequence of commands

Given that each verb in dplyr specializes in a particular task, addressing complex queries often typically involves integrating several verbs, a process simplified by using the native forward pipe operator, `|>`.

For example, we want to filter the original dataset to include only rows where the shock index category is “low” or “high”, using a sequence of dplyr verbs.

```{r}
sub_dat <- dat |>
  mutate(shock_index = heart_rate / sbp,
         shock_index1 = case_when(shock_index < 0.5 ~ "low",
                                  shock_index <= 0.7 ~ "normal",
                                  TRUE ~ "high")) |> 
  filter(shock_index1 == "low" | shock_index1 == "high")

```

<br>

## Plots

-   **Basic plot**

```{r}
# Basic plot
plot(dat10$age, dat10$shock_index,
     pch = 19,               # Point shape (19 is a solid circle)
     cex = 1.5,              # Point size
     xlab = "Age (years)",   # X-axis label
     ylab = "Shock Index",   # Y-axis label
     main = "Age vs Shock Index", # Optional title
     las = 1,                # Makes axis numbers horizontal (easier to read)
     frame.plot = FALSE)     # Removes the top/right border for a cleaner look
```

<br>

-   **ggplot**

We can also use the `ggplot()` function from the `ggplot2` package to create a similar plot. Notably, `ggplot2` is one of the core packages included in the `tidyverse`, so it is loaded into an R session when we run the command `library(tidyverse)`.

```{r}
ggplot(dat10, aes(x = age, y = shock_index)) +
  geom_point(size = 2.5) +
  labs(title = "Age vs Shock Index",
    x = "Age (years)",
    y = "Shock Index"
  )
```


- `aes()` stands for aesthetic mappings. It defines how variables from the dataset are mapped to visual properties of the plot, such as x-axis position and y-axis position.

- `geom_point()` is a geometry function that draws points on the plot.


<br>


## Reshaping data

Reshaping data between “wide” and “long” formats in R is a common task in data transformation. For example, data were collected from four individuals, with total cholesterol (TCH) levels measured at baseline (week 0), as well as at 6 and 12 weeks following a 12-week dietary intervention. The data are stored in an Excel spreadsheet with the following structure:

```{r}
dat_TCH <- read_excel("data/cholesterol.xlsx")
dat_TCH
```

We observe that the dataset is organized in what is commonly referred to as a “wide” format. This structure is particularly well-suited for a repeated measures design, also known as a longitudinal or within-subject design. In such designs, the same individuals are measured multiple times over a period, with each measurement representing a different time point.

<br> First, we add a column to the dataset that assigns a unique identifier to each row using the `rowid_to_column()` function:

```{r}
dat_TCH <- dat_TCH |> 
  rowid_to_column()
dat_TCH
```

-   **From wide to long format**

```{r}
dat_TCH_long <- dat_TCH |> 
  pivot_longer(cols = starts_with("week"),
               names_to = "week",
               values_to = "total_cholesterol")
dat_TCH_long
```

-   **From long to wide format**

```{r}
dat_TCH_wide <- dat_TCH_long |>
  pivot_wider(id_cols = rowid,
              names_from = week,
              values_from = total_cholesterol)
dat_TCH_wide
```
